---
title: Introduction to `duckdb` and SQL for register-scale data work
date: last-modified
toc: true
numbersections: true
mainfont: Times New Roman
monofont: Times New Roman
sansfont: Times New Roman
seriffont: Times New Roman
highlight-style: tango
fontsize: 10pt
geometry:
    - margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE
    )
```

\clearpage
# Introduction

This document is a hands-on companion to the [`RA guide`](https://github.com/EriksenJ/ra_guide) section on larger-than-memory datasets. It shows how to use the [`duckdb`](https://duckdb.org/) package from R to query parquet files with SQL, without loading the data into memory.

The example data is simulated to look like Danish registers (BEF for demographics, IND for annual income), using canonical DST variable names (`PNR`, `AAR`, `ALDER`, `PERINDKIALT13`). You can run everything in this document without access to DST — the setup step below generates the parquet files locally.

The recommended external reading is Grant McDermott's [Data wrangling with DuckDB](https://grantmcdermott.com/duckdb-polars/duckdb-sql.html), which covers both the dplyr and SQL modes side by side.


# Why SQL for register work?

You can drive `duckdb` from R in two ways:

1. **dplyr verbs** via `to_duckdb()` (from the `arrow` package) plus `dbplyr`. Familiar if you already use `dplyr`; each verb chain gets translated into SQL under the hood.
2. **Plain SQL** via `dbGetQuery()` or `dbSendQuery()`. You write the SQL yourself.

For simple one-table filters and aggregations, dplyr-in-duckdb is fine and often the fastest way to get an answer. For anything more involved — joins across three or more registers, window functions (lags, running totals, ranks), CTEs to structure a multi-step query, or features `dbplyr` does not translate — SQL is usually clearer, faster to iterate on, and gives you the full duckdb feature set.

The rule of thumb: **do the heavy lifting on disk in duckdb, `collect()` the small aggregated result into R for modeling and plotting.**


# Setup

```{r}
pacman::p_load(
    duckdb,
    dplyr,
    arrow,
    rio,
    here
)
```

We will create a small working directory for the simulated parquet files.

```{r}
data_dir = here::here("duckdb_intro", "sim_data")
dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
```


# Simulate register-style data

We simulate two datasets:

- `bef.parquet`: one row per person per year, with age (`ALDER`).
- `ind.parquet`: one row per person per year, with personal income (`PERINDKIALT13`).

Both are keyed on `(PNR, AAR)`.

```{r}
set.seed(42)

n_pnr = 5000
years = 2008:2022

pnr_pool = sprintf("%010d", sample.int(1e9, n_pnr))
birth_yr = sample(1950:2000, n_pnr, replace = TRUE)

bef = expand.grid(PNR = pnr_pool, AAR = years, stringsAsFactors = FALSE) |>
    as.data.frame()
bef$ALDER = bef$AAR - birth_yr[match(bef$PNR, pnr_pool)]
bef = subset(bef, ALDER >= 0 & ALDER <= 100)

ind = bef[, c("PNR", "AAR")]
ind$PERINDKIALT13 = round(
    exp(rnorm(nrow(ind), mean = 12.5, sd = 0.6)) *
        (1 + 0.01 * (bef$ALDER - 40))
)

bef |> export(file.path(data_dir, "bef.parquet"))
ind |> export(file.path(data_dir, "ind.parquet"))

list.files(data_dir)
```


# Connecting to duckdb

Open an in-memory connection. Passing no path keeps everything in RAM; pass a file path (e.g. `"my_project.duckdb"`) if you want a persistent database. For querying parquet files, in-memory is usually all you need.

```{r}
con = dbConnect(duckdb())
```


# Querying a parquet file

DuckDB reads parquet files directly — you do not have to load them into a table first. Just reference the path in `FROM`.

```{r}
dbGetQuery(con, sprintf(
    "SELECT COUNT(*) AS n_rows FROM '%s'",
    file.path(data_dir, "bef.parquet")
))
```

For readability we can register the paths as duckdb views once, then reference them by name for the rest of the session:

```{r}
dbExecute(con, sprintf(
    "CREATE OR REPLACE VIEW bef AS SELECT * FROM '%s'",
    file.path(data_dir, "bef.parquet")
))
dbExecute(con, sprintf(
    "CREATE OR REPLACE VIEW ind AS SELECT * FROM '%s'",
    file.path(data_dir, "ind.parquet")
))
```

Now the queries below can just say `FROM bef` and `FROM ind`.


# Filter and aggregate

Mean income by year, restricted to working-age respondents:

```{r}
dbGetQuery(con, "
  SELECT AAR,
         AVG(PERINDKIALT13) AS mean_income,
         COUNT(*)           AS n_obs
  FROM ind
  JOIN bef USING (PNR, AAR)
  WHERE ALDER BETWEEN 25 AND 60
  GROUP BY AAR
  ORDER BY AAR
")
```


# Join across registers, group, and lag with a window function

The realistic register task: join demographics and income on `(PNR, AAR)`, restrict to a sample, aggregate by age × year, and add each age group's prior-year mean using `LAG(...) OVER (PARTITION BY ALDER ORDER BY AAR)`.

```{r}
query = "
  WITH joined AS (
    SELECT b.PNR, b.AAR, b.ALDER, i.PERINDKIALT13
    FROM bef b
    JOIN ind i USING (PNR, AAR)
    WHERE b.ALDER BETWEEN 25 AND 60
      AND b.AAR   BETWEEN 2010 AND 2020
  )
  SELECT
    AAR,
    ALDER,
    AVG(PERINDKIALT13)                                             AS mean_income,
    LAG(AVG(PERINDKIALT13)) OVER (PARTITION BY ALDER ORDER BY AAR) AS mean_income_lag1,
    COUNT(*)                                                       AS n_obs
  FROM joined
  GROUP BY AAR, ALDER
  ORDER BY ALDER, AAR
"

result = dbGetQuery(con, query)
head(result, 12)
```

The result is a plain `data.frame`, ready to save or plot:

```{r}
result |> export(file.path(data_dir, "mean_income_by_alder_year.parquet"))
```


# The same query with dplyr verbs

For contrast, here is the aggregation (without the lag — see below) written with dplyr on a `to_duckdb()` connection:

```{r}
bef_tbl = open_dataset(file.path(data_dir, "bef.parquet")) |> to_duckdb(con = con)
ind_tbl = open_dataset(file.path(data_dir, "ind.parquet")) |> to_duckdb(con = con)

bef_tbl |>
    inner_join(ind_tbl, by = c("PNR", "AAR")) |>
    filter(ALDER >= 25, ALDER <= 60, AAR >= 2010, AAR <= 2020) |>
    summarise(
        mean_income = mean(PERINDKIALT13, na.rm = TRUE),
        n_obs       = n(),
        .by = c(AAR, ALDER)
    ) |>
    arrange(ALDER, AAR) |>
    head(12) |>
    collect()
```

`dbplyr` does not translate window functions like `LAG(...) OVER (...)` in a way that stays inside duckdb — you would have to `collect()` and then compute the lag in R. That is fine for this small result, but for a wider register sample where you want the lag computed before `collect()`, SQL is the right tool.


# Inspect what dplyr sends to duckdb

You can see the SQL that dplyr generates:

```{r}
bef_tbl |>
    filter(ALDER >= 25, ALDER <= 60) |>
    summarise(mean_alder = mean(ALDER), .by = AAR) |>
    show_query()
```


# When to use what

Our default is **SQL for anything on-disk, R for anything in-memory** (see the [`RA guide`](https://github.com/EriksenJ/ra_guide) for the reasoning):

- **SQL in duckdb** for any transformation on a register-sized parquet file — joins, aggregations, window functions, sample construction.
- **dplyr on `to_duckdb()`** only for exploration — poking at a dataset to figure out what SQL you eventually want to write. Once the query becomes pipeline code, port it to SQL. The `show_query()` trick above is a useful stepping stone if you are still learning SQL.
- **Base R / `data.table` / `fixest` / `ggplot2`** for everything downstream of `collect()` — modeling, plotting, small-table manipulation.


# Cleanup

```{r}
dbDisconnect(con, shutdown = TRUE)
```


# References

- Grant McDermott, [Data wrangling with DuckDB](https://grantmcdermott.com/duckdb-polars/duckdb-sql.html).
- DuckDB [SQL introduction](https://duckdb.org/docs/sql/introduction) and [window functions](https://duckdb.org/docs/sql/window_functions).
- [awesome-arrow-r](https://github.com/thisisnic/awesome-arrow-r) — the `arrow` + `duckdb` combination in R.


# TODO 

- Add an introduction to distributed work on the DST server - based on Anders Humlum's [guide](https://www.andershumlum.com/codes)